---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 2 - Due date 02/03/23"
author: "John Rooney"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2022 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
#Importing data set
energy_data <- read.csv(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv", header=TRUE,dec = ".",sep=",",stringsAsFactors = TRUE)
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
energy_data_df<-energy_data %>%
  select(Month, Total.Biomass.Energy.Production, Total.Renewable.Energy.Production, Hydroelectric.Power.Consumption)
head(energy_data_df)
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
ts_energy_data<-ts(energy_data_df[,2:4], frequency=12,start=c(1973,1))
head(ts_energy_data)
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
mean(energy_data_df$Total.Biomass.Energy.Production)
sd(energy_data_df$Total.Biomass.Energy.Production)
mean(energy_data_df$Total.Renewable.Energy.Production)
sd(energy_data_df$Total.Renewable.Energy.Production)
mean(energy_data_df$Hydroelectric.Power.Consumption)
sd(energy_data_df$Hydroelectric.Power.Consumption)
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
plot(ts_energy_data[,"Total.Biomass.Energy.Production"],type="l",col="blue",xlab="Year",ylab="Trillion Btu",main="Total Biomass Energy Production") 
abline(h=mean(ts_energy_data[,"Total.Biomass.Energy.Production"]),col="red")

plot(ts_energy_data[,"Total.Renewable.Energy.Production"],type="l",col="green",xlab="Year",ylab="Trillion Btu",main="Total Renewable Energy Production")
abline(h=mean(ts_energy_data[,"Total.Renewable.Energy.Production"]),col="red")

plot(ts_energy_data[,"Hydroelectric.Power.Consumption"],type="l",col="orange",xlab="Year",ylab="Trillion Btu",main="Hydroelectric Power Consumption")
abline(h=mean(ts_energy_data[,"Hydroelectric.Power.Consumption"]),col="red")
```
Total biomass energy production grew rapidly through the mid-1980s, then stayed relatively stable through 2000 when it again grew rapidly up through 2015. Total renewable energy was relatively stable from 1980 through 2000 when it began to grow rapidly. Total hydroelectric power looks highly seasonal, showing decreases starting around 2000 which would appear to correlate with production of biomass and renewable energy. 

## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor(ts_energy_data)
```
Total biomass energy production and total renewable energy production are strongly positively correlated, with values of 1 to .92 and vice versa. Each is slightly negatively correlated with hydroelectric power consumption, with hydroelectric power consumption to biomass energy production having a correlation value of -0.3 and hydroelectric power consumption to renewable energy production having a correlation value of -0.1. 

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
Acf(ts_energy_data[,1], lag.max = 40, main="Total Biomass Energy Production", ylim=c(-1,1))
Acf(ts_energy_data[,2], lag.max = 40, main="Total Renewable Energy Production", ylim=c(-1,1))
Acf(ts_energy_data[,3], lag.max = 40, main="Total Hydroelectric Power Consumption", ylim=c(-1,1))
```
Total biomass energy production and total renewable energy production show similar behavior, both with quite high significance that slowly decreases over time. Total hydroelectric power consumption shows more seasonality, with higher significance at the beginning and end of each year. This significance decreases over time. 

## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
Pacf(ts_energy_data[,1], lag.max = 40, main="Total Biomass Energy Production", ylim=c(-1,1))
Pacf(ts_energy_data[,2], lag.max = 40, main="Total Renewable Energy Production", ylim=c(-1,1))
Pacf(ts_energy_data[,3], lag.max = 40, main="Total Hydroelectric Power Consumption", ylim=c(-1,1))
```
Total biomass energy production and renewable energy production again show similar behavior, although far fewer data points are significant. In both cases significant moments in time are right at the beginning and end of the year. In both cases significance decreases over time, especially so for total renewable energy production. For total hydroelectric power, data appears significant towards the end of the first first and far less in further years, with significant data showing only at the beginning of each year. 

